How is experience transformed into memory?

### 2018-09-02 paper/main.tex
additional edits to results, got down some discussion thoughts in comments

Visual inspection of the video and recall topic trajectories reveals a striking pattern.  First, the topic trajectory of the video (which reflects its dynamic content; Fig.~\ref{fig:trajectory}A) is captured nearly perfectly by the averaged topic trajectories of participants' recalls (Fig.~\ref{fig:trajectory}B).  To assess the consistency of these recall trajectories across participants, we asked: given that a participant's recall trajectory had entered a particular location in topic space, could the position of their \textit{next} recalled event be predicted reliably?  For each location in topic space, we computed the set of line segments connecting successively recalled events (across all participants) that intersected that location (see \textit{Methods} for additional details).  We then computed (for each location) the distribution of angles formed by the lines defined by those line segments and a fixed reference line (the $x$-axis).  Rayleigh tests revealed the set of locations in topic space at which these across-participant distributions exhibited reliable peaks (blue arrows in Fig.~\ref{fig:trajectory}B).  We observed that the locations traversed by nearly the entire the video trajectory exhibited such peaks.  In other words, participants exhibited similar trajectories that also matched the trajectory of the original video (Fig.~\ref{fig:trajectory}C).  This is especially notable when considering the fact that the number of events participants recalled (dots in Fig.~\ref{fig:trajectory}C) varied considerably across people, and that every participant used different words to describe what they had remembered happening in the video.

@@

\section*{Discussion}
\label{sec:discussion}
Studying human memory is commonly distilled down to a process of matching specific moments of a past experience with specific mnemonic outcomes.  In traditional trial-based free recall experiments, individual stimuli encountered during encoding are typically labeled as ``remembered'' or ``forgotten'' depending on whether the stimulus is recalled/recognized during a subsequent test. While this approach has advanced our understanding of human memory immensely, it does not translate well to naturalistic experiences. For one thing, the contents of our recall at any given moment might be reflected in many prior experiences/moments. Furthermore, the particular words used to describe the experience will inevitably vary across people and even across repeated recollections within an individual. Thus, there is not a ``one-to-one'' mapping between naturalistic experiences and their mnemonic counterparts. For example, remembering the patterns and colors of each person's shirt in a crowd might be considered as excellent recall in a standard memory task setup. But if the rememberer failed to note that the people in the crowd were gathered for their surprise birthday party, then they would have missed the "point" of the experience.

>>>>

Visual inspection of the video and recall topic trajectories reveals a striking pattern.  First, the topic trajectory of the video (which reflects its dynamic content; Fig.~\ref{fig:trajectory}A) is captured nearly perfectly by the averaged topic trajectories of participants' recalls (Fig.~\ref{fig:trajectory}B).  To assess the consistency of these recall trajectories across participants, we asked: given that a participant's recall trajectory had entered a particular location in topic space, could the position of their \textit{next} recalled event be predicted reliably?  For each location in topic space, we computed the set of line segments connecting successively recalled events (across all participants) that intersected that location (see \textit{Methods} for additional details).  We then computed (for each location) the distribution of angles formed by the lines defined by those line segments and a fixed reference line (the $x$-axis).  Rayleigh tests revealed the set of locations in topic space at which these across-participant distributions exhibited reliable peaks (blue arrows in Fig.~\ref{fig:trajectory}B).  We observed that the locations traversed by nearly the entire the video trajectory exhibited such peaks.  In other words, participants exhibited similar trajectories that also matched the trajectory of the original video (Fig.~\ref{fig:trajectory}C).  This is especially notable when considering the fact that the number of events participants recalled (dots in Fig.~\ref{fig:trajectory}C) varied considerably across people, and that every participant used different words to describe what they had remembered happening in the video.  Differences in the numbers of remembered events appear in participants' trajectories as differences in the sampling resolution along the trajectory.  We note that this framework also provides a means of detangling classic ``proportion recalled'' measures (i.e., the proportion of video events referenced in participants' recalls) from participants' abilities to recaputulate the full gist of the original video (i.e., the similarity in the shape of the original video trajectory and that defined by each participant's reocounting of the video).

@@

\section*{Discussion}
\label{sec:discussion}
% JRM THOUGHTS ON DISCUSSION:
% - Nature says "2-3 paragraphs" -- needs to be much shorter.
% - I think we can get rid of anything we mention in the intro or results
% - Things not easily presented elsewhere are the background on the brain results and applications to education metrics.  I think we can focus on those two parts, add an "overview" and "conclusion" (potentially), and cut the rest.
Studying human memory is commonly distilled down to a process of matching specific moments of a past experience with specific mnemonic outcomes.  In traditional trial-based free recall experiments, individual stimuli encountered during encoding are typically labeled as ``remembered'' or ``forgotten'' depending on whether the stimulus is recalled/recognized during a subsequent test. While this approach has advanced our understanding of human memory immensely, it does not translate well to naturalistic experiences. For one thing, the contents of our recall at any given moment might be reflected in many prior experiences/moments. Furthermore, the particular words used to describe the experience will inevitably vary across people and even across repeated recollections within an individual. Thus, there is not a ``one-to-one'' mapping between naturalistic experiences and their mnemonic counterparts. For example, remembering the patterns and colors of each person's shirt in a crowd might be considered as excellent recall in a standard memory task setup. But if the rememberer failed to note that the people in the crowd were gathered for their surprise birthday party, then they would have missed the "point" of the experience.
